\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-29 paper.qmd
Correct reference rendering issue in Introduction section

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional ($k-D$) representation of high-dimensional ($p-D$) data. Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008], uniform manifold approximation and projection (UMAP) [@Leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@article03], large-scale dimensionality reduction Using triplets (TriMAP) [@article02], and pairwise controlled manifold approximation (PaCMAP) [@Yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in $p-D$. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the $k-D$ minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@article64] spreading to capture geometric shapes, that include both global and local structure.

@@

The alternative approach to visualizing the high-dimensional data is to use linear projections. PCA is the classical approach, resulting in a set of new variables which are linear combinations of the original variables. Tours, defined by @Asimov1985, broaden the scope by providing movies of linear projections, that provide views the data from all directions. @lee2021review provides an review of the main developments in tours. There are many tour algorithms implemented, with many available in the R package `tourr` [@wickham2011], and versions enabling better interactivity in `langevitour` [@harisson2024] and `detourr` [@hart2022]. Linear projections are a safe way to view high-dimensional data, because they do not warp the space, so they are more faithful representations of the structure. 
However, linear projections can be cluttered, and global patterns can obscure local structure. The simple activity of projecting data from $p-D$ suffers from piling [@laa2022], where data concentrates in the center of projections. NLDR is designed to escape these issues, to exaggerate structure so that it can be observed. But as a result NLDR can hallucinate wildly, to suggest patterns that are not actually present in the data.

>>>>

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional ($k-D$) representation of high-dimensional ($p-D$) data. Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2022], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in $p-D$. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the $k-D$ minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

The alternative approach to visualizing the high-dimensional data is to use linear projections. PCA is the classical approach, resulting in a set of new variables which are linear combinations of the original variables. Tours, defined by @asimov1985, broaden the scope by providing movies of linear projections, that provide views the data from all directions. @lee2021 provides an review of the main developments in tours. There are many tour algorithms implemented, with many available in the R package `tourr` [@wickham2011], and versions enabling better interactivity in `langevitour` [@harisson2024] and `detourr` [@hart2022]. Linear projections are a safe way to view high-dimensional data, because they do not warp the space, so they are more faithful representations of the structure. 
However, linear projections can be cluttered, and global patterns can obscure local structure. The simple activity of projecting data from $p-D$ suffers from piling [@laa2022], where data concentrates in the center of projections. NLDR is designed to escape these issues, to exaggerate structure so that it can be observed. But as a result NLDR can hallucinate wildly, to suggest patterns that are not actually present in the data.

### 2024-03-29 paper.tex
Correct reference rendering issue in Introduction section

Non-linear dimension reduction (NLDR) is popular for making a convenient
low-dimensional (\(k-D\)) representation of high-dimensional (\(p-D\))
data. Recently developed methods include t-distributed stochastic
neighbor embedding (tSNE) \citep{Laurens2008}, uniform manifold
approximation and projection (UMAP) \citep{Leland2018}, potential of
heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm
\citep{article03}, large-scale dimensionality reduction Using triplets
(TriMAP) \citep{article02}, and pairwise controlled manifold
approximation (PaCMAP) \citep{Yingfan2021}. However, the representation
generated can vary dramatically from method to method, and with
different choices of parameters or random seeds made using the same
method (Figure~\ref{fig-NLDR-variety}). The dilemma for the analyst is
then, \textbf{which representation to use}. The choice might result in
different procedures used in the downstream analysis, or different
inferential conclusions. The research described here provides new visual
tools to aid with this decision.

@@

Many new methods for NLDR have emerged in recent years, all designed to
better capture specific structures potentially existing in \(p-D\). Here
we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP
and PaCMAP. tNSE and UMAP can be considered to produce the \(k-D\)
minimizing the divergence between two distributions, where the
distributions are modeling the inter-point distances. PHATE, TriMAP and
PaCMAP are examples of diffusion processes \citep{article64} spreading
to capture geometric shapes, that include both global and local
structure.

@@

The alternative approach to visualizing the high-dimensional data is to
use linear projections. PCA is the classical approach, resulting in a
set of new variables which are linear combinations of the original
variables. Tours, defined by \citet{Asimov1985}, broaden the scope by
providing movies of linear projections, that provide views the data from
all directions. \citet{lee2021review} provides an review of the main
developments in tours. There are many tour algorithms implemented, with
many available in the R package \texttt{tourr} \citep{wickham2011}, and
versions enabling better interactivity in \texttt{langevitour}
\citep{harisson2024} and \texttt{detourr} \citep{hart2022}. Linear
projections are a safe way to view high-dimensional data, because they
do not warp the space, so they are more faithful representations of the
structure. However, linear projections can be cluttered, and global
patterns can obscure local structure. The simple activity of projecting
data from \(p-D\) suffers from piling \citep{laa2022}, where data
concentrates in the center of projections. NLDR is designed to escape
these issues, to exaggerate structure so that it can be observed. But as
a result NLDR can hallucinate wildly, to suggest patterns that are not
actually present in the data.

>>>>

Non-linear dimension reduction (NLDR) is popular for making a convenient
low-dimensional (\(k-D\)) representation of high-dimensional (\(p-D\))
data. Recently developed methods include t-distributed stochastic
neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold
approximation and projection (UMAP) \citep{leland2018}, potential of
heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm
\citep{moon2019}, large-scale dimensionality reduction Using triplets
(TriMAP) \citep{amid2022}, and pairwise controlled manifold
approximation (PaCMAP) \citep{yingfan2021}. However, the representation
generated can vary dramatically from method to method, and with
different choices of parameters or random seeds made using the same
method (Figure~\ref{fig-NLDR-variety}). The dilemma for the analyst is
then, \textbf{which representation to use}. The choice might result in
different procedures used in the downstream analysis, or different
inferential conclusions. The research described here provides new visual
tools to aid with this decision.

@@

Many new methods for NLDR have emerged in recent years, all designed to
better capture specific structures potentially existing in \(p-D\). Here
we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP
and PaCMAP. tNSE and UMAP can be considered to produce the \(k-D\)
minimizing the divergence between two distributions, where the
distributions are modeling the inter-point distances. PHATE, TriMAP and
PaCMAP are examples of diffusion processes \citep{coifman2005} spreading
to capture geometric shapes, that include both global and local
structure.

@@

The alternative approach to visualizing the high-dimensional data is to
use linear projections. PCA is the classical approach, resulting in a
set of new variables which are linear combinations of the original
variables. Tours, defined by \citet{asimov1985}, broaden the scope by
providing movies of linear projections, that provide views the data from
all directions. \citet{lee2021} provides an review of the main
developments in tours. There are many tour algorithms implemented, with
many available in the R package \texttt{tourr} \citep{wickham2011}, and
versions enabling better interactivity in \texttt{langevitour}
\citep{harisson2024} and \texttt{detourr} \citep{hart2022}. Linear
projections are a safe way to view high-dimensional data, because they
do not warp the space, so they are more faithful representations of the
structure. However, linear projections can be cluttered, and global
patterns can obscure local structure. The simple activity of projecting
data from \(p-D\) suffers from piling \citep{laa2022}, where data
concentrates in the center of projections. NLDR is designed to escape
these issues, to exaggerate structure so that it can be observed. But as
a result NLDR can hallucinate wildly, to suggest patterns that are not
actually present in the data.
